### 2023-04-26 [2020-08]_ACMTOMS_(MQSI)/paper/LaTeX/4-analysis.tex

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
{\everymath={\scriptstyle}

\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
  facet model's sensitivity to small data perturbations. This example is
  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have
  equal second derivatives. Given the exact five data points
  seen above, the quadratic facet model produces the slope seen in the
  solid blue line at $x = 5/2$. However, by subtracting the value of
  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
  quadratic facet model produces the slope seen in the dashed red line
  at $x = 5/2$. This is the nature of a facet model and a side effect
  of associating data with local facets.
\par} \endinsert

\heading{4. COMPLEXITY AND SENSITIVITY}

Algorithms 1 and 4 have ${\cal O}(n)$ runtime for $n$ data
points. Algorithm 2 has a fixed cost ${\cal O}(1)$. Given a fixed
schedule for shrinking derivative values, Algorithm 3 has a ${\cal
  O}(n)$ runtime for $n$ data points. In execution, the majority of
the time, still ${\cal O}(n)$, is spent solving the banded linear
system of equations for the B-spline coefficients. Thus for $n$ data
points, the overall execution time is ${\cal O}(n)$.

@@

The binary search procedure provably converges on the boundary of the
region of monotonicity precisely defined by Ulrich and Watson [1994]
through the application of a Boolean function that is guaranteed to be
true at one end of an interval. Since a local quadratic interpolant is
applied for initial function derivative estimates, the approximation
order of the resulting fit is $O(h^3)$ (as for any second order
approximation), but it should be noted that asymptotic approximation
order is rarely of importance when considering the scattered sparse
data that $C^2$ approximations like {\tt MQSI} provide. Were exact
evaluations provided, higher order methods always win (all
other things being equal) as the data density increases. For sparse
data, what constitutes a better fit is either subjective or dependent
on the problem.

>>>>

\begin{figure}
\includegraphics[width=4in]{vis/1-sensitivity.eps}
\caption{ % \everymath={\scriptstyle}  % \narrower\noindent\rmVIII Fig.\ 1.
  A demonstration of the quadratic facet model's sensitivity to small
  data perturbations. This example is composed of two quadratic
  functions $f_1(x) = x^2$ over points $\{1$, $2$, $5/2\}$, and
  $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$, $4\}$. Notably,
  $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have equal second
  derivatives. Given the exact five data points seen above, the
  quadratic facet model produces the slope seen in the solid blue line
  at $x = 5/2$. However, by subtracting the value of $f_3$ $=
  \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$ is the
  machine precision ($2^{-52}$ for an IEEE 64-bit real), the quadratic
  facet model produces the slope seen in the dashed red line at $x =
  5/2$. This is the nature of a facet model and a side effect of
  associating data with local facets.}
\end{figure}

\section{COMPLEXITY AND SENSITIVITY}

Algorithms 1 and 4 have $\mathcal{O}(n)$ runtime for $n$ data
points. Algorithm 2 has a fixed cost $\mathcal{O}(1)$. Given a fixed
schedule for shrinking derivative values, Algorithm 3 has a
$\mathcal{O}(n)$ runtime for $n$ data points. In execution, the
majority of the time, still $\mathcal{O}(n)$, is spent solving the banded
linear system of equations for the B-spline coefficients. Thus for $n$
data points, the overall execution time is $\mathcal{O}(n)$.

@@

The binary search procedure provably converges on the boundary of the
region of monotonicity precisely defined by Ulrich and Watson
\cite{ulrich1994positivity} through the application of a Boolean
function that is guaranteed to be true at one end of an
interval. Since a local quadratic interpolant is applied for initial
function derivative estimates, the approximation order of the
resulting fit is $O(h^3)$ (as for any second order approximation), but
it should be noted that asymptotic approximation order is rarely of
importance when considering the scattered sparse data that $C^2$
approximations like {\tt MQSI} provide. Were exact evaluations
provided, higher order methods always win (all other things being
equal) as the data density increases. For sparse data, what
constitutes a better fit is either subjective or dependent on the
problem.

### 2023-04-26 [2020-08]_ACMTOMS_(MQSI)/paper/TeX/1-introduction.tex

It is often the case that the true underlying function or phenomenon
being modeled has known properties like convexity, positivity,
various levels of continuity, or monotonicity. Given a reasonable
amount of data, it quickly becomes difficult to achieve desirable
properties in a single polynomial function. In general, the
maintenance of function properties through interpolation/regression is
referred to as {\it shape preserving} [Fritsch and Carlson 1980;
Gregory 1985]. The specific properties the present algorithm will
preserve in approximations are monotonicity and $C^2$ continuity. 
\beginred
Notably this work does not consider convexity preservation, as Mulansky
and Neamtu [1991] prove that there exists strictly convex data sets which
do not allow convex interpolation with $C^2$ splines of fixed degree.
\endred
In addition to previously mentioned applications, \red{being $C^2$ and monotonicity preserving is}
crucially important in statistics to the approximation of a cumulative
distribution function and subsequently the effective generation of
random numbers from a specified distribution [Ramsay 1988].  A spline
function with these properties could approximate a cumulative
distribution function to a high level of accuracy with relatively few
intervals. A twice continuously differentiable approximation to a
cumulative distribution function (CDF) would produce a corresponding
probability density function (PDF) that is continuously
differentiable, which is desirable.

\beginred

@@

\endred

The theory for \red{{\it sharp}} necessary and sufficient bounds for
monotone $C^2$ quintic interpolation has been provided [Ulrich and
  Watson 1990; 1994] and that theory was recently utilized in a
proposed algorithm [Lux et al. 2020; Lux 2020] for monotone $C^2$
quintic spline construction, however no published mathematical
software exists for the quintic case \red{based on sharp monotonicity
conditions}. The software presented here represents the first published
software package for producing $C^2$ shape preserving splines
\red{based on sharp monotonicity conditions}. This work improves upon
the algorithm presented by Lux et\ al.\ [2020] by refactoring
computations for improved numerical stability, estimating minimum
magnitude second derivatives at breakpoints with a quadratic facet
model, and using a binary search to reduce the magnitude of the
modifications made to initial derivative estimates when constructing a
monotone spline interpolant.

\beginred

\heading{Overview}

\endred

This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
sharp necessary and sufficient conditions in Ulrich and Watson [1994]
for the construction of monotone $C^2$ quintic spline interpolants of
monotone data. Precisely, the problem is, given a strictly increasing
sequence $X_1<X_2< \cdots <X_n$ of breakpoints with corresponding
monotone increasing function values $Y_1\le Y_2\le \cdots \le Y_n$,
find a $C^2$ monotone increasing quintic spline $Q(x)$ with the same
breakpoints satisfying $Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI}
actually does something slightly more general, producing $Q(x)$ that
is monotone increasing (decreasing) wherever the data is monotone
increasing (decreasing).)

The remainder of this paper is structured as follows: Section 2
provides the algorithms for constructing a monotone $C^2$ quintic
spline interpolant to monotone data, Section 3 outlines the method of
spline representation ($B$-spline basis) and evaluation, Section 4
analyzes the complexity and sensitivity of the algorithms in {\tt
  MQSI}, and Section 5 presents timing data and some graphs of
constructed interpolants \red{as well as a visual comparison with
  existing monotone spline packages}.

>>>>

It is often the case that the true underlying function or phenomenon
being modeled has known properties like convexity, positivity,
various levels of continuity, or monotonicity. Given a reasonable
amount of data, it quickly becomes difficult to achieve desirable
properties in a single polynomial function. In general, the
maintenance of function properties through interpolation/regression is
referred to as {\it shape preserving} [Fritsch and Carlson 1980;
Gregory 1985]. The specific properties the present algorithm will
preserve in approximations are monotonicity and $C^2$ continuity. 
Notably this work does not consider convexity preservation, as Mulansky
and Neamtu [1991] prove that there exists strictly convex data sets which
do not allow convex interpolation with $C^2$ splines of fixed degree.
In addition to previously mentioned applications, being $C^2$ and monotonicity preserving is
crucially important in statistics to the approximation of a cumulative
distribution function and subsequently the effective generation of
random numbers from a specified distribution [Ramsay 1988].  A spline
function with these properties could approximate a cumulative
distribution function to a high level of accuracy with relatively few
intervals. A twice continuously differentiable approximation to a
cumulative distribution function (CDF) would produce a corresponding
probability density function (PDF) that is continuously
differentiable, which is desirable.

@@

The theory for {\it sharp} necessary and sufficient bounds for
monotone $C^2$ quintic interpolation has been provided [Ulrich and
  Watson 1990; 1994] and that theory was recently utilized in a
proposed algorithm [Lux et al. 2020; Lux 2020] for monotone $C^2$
quintic spline construction, however no published mathematical
software exists for the quintic case based on sharp monotonicity
conditions. The software presented here represents the first published
software package for producing $C^2$ shape preserving splines
based on sharp monotonicity conditions. This work improves upon
the algorithm presented by Lux et\ al.\ [2020] by refactoring
computations for improved numerical stability, estimating minimum
magnitude second derivatives at breakpoints with a quadratic facet
model, and using a binary search to reduce the magnitude of the
modifications made to initial derivative estimates when constructing a
monotone spline interpolant.

\heading{Overview}
This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
sharp necessary and sufficient conditions in Ulrich and Watson [1994]
for the construction of monotone $C^2$ quintic spline interpolants of
monotone data. Precisely, the problem is, given a strictly increasing
sequence $X_1<X_2< \cdots <X_n$ of breakpoints with corresponding
monotone increasing function values $Y_1\le Y_2\le \cdots \le Y_n$,
find a $C^2$ monotone increasing quintic spline $Q(x)$ with the same
breakpoints satisfying $Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI}
actually does something slightly more general, producing $Q(x)$ that
is monotone increasing (decreasing) wherever the data is monotone
increasing (decreasing).)

The remainder of this paper is structured as follows: Section 2
provides the algorithms for constructing a monotone $C^2$ quintic
spline interpolant to monotone data, Section 3 outlines the method of
spline representation ($B$-spline basis) and evaluation, Section 4
analyzes the complexity and sensitivity of the algorithms in {\tt
  MQSI}, and Section 5 presents timing data and some graphs of
constructed interpolants as well as a visual comparison with
  existing monotone spline packages.
